### 2024-06-11 manuscript.Rmd
added even more text

---
title             : "`reprodubileRchunks`: R Markdown Code Chunks with Testable Reproducibility"
shorttitle        : "Title"

@@

abstract: |
 Research is considered _reproducible_ if the same analysis on the same data yields the same results if performed on a different computer or on the same computer later in time. Reproducibility is a prerequisite for replicable, robust and transparent research.
 Various approaches have been suggested to increase chances of reproducibility.
 Many of them rely on R Markdown as a language to dynamically generate reproducible research assets (e.g., reports, posters, presentation).
 However, a simple way to test whether different parts of these assets are reproducible is still missing.
 We introduce the R package `reproducibleRchunks`, which provides a new type of code chunk in R Markdown documents, which automatically store meta data about original computational results and check later reproductions of the research assets for reproduction. 
 With a minimal change to users' workflows, we hope that this approach increases transparency and trustworthiness of research assets.

keywords          : "reproducibility, Markdown, R"
wordcount         : "X"

@@

R Markdown is a simple language, which allows mixing natural language, simple formatting instructions (e.g., what is a headline, what should appear in bold face, what are list elements) and computer code (R, python or others) in a single document [@xie2018r].
When ever the document is compiled (or, _knitted_ in R speak), the computer code chunks are executed and allow for the dynamic generation of parts of the document.
A particular advantage of R Markdown is that it helps avoiding common threats to reproducibility of code, first and foremost, copy&paste errors between results of a regular R script and a scientific report (e.g., a Word document) [@peikert2021reproducible];
further, it is suitable to generate a variety of scientific assets such as posters [@posterdown2019], presentations [@xaringan2022], resumés [@vitae2023] or preregistrations [@peikert2021reproducible] from the same Markdown language.

The `reproducibleRchunk` package allows you to make computational results in R automatically testable for reproduction (does the same script with the same data produce the same results, e.g. on a different computer and/or later in time) with minimal changes to your workflow.
There is only a single thing you need to change in your analysis if you are already using R Markdown: In the first code chunk of your document, load the package and then change the code chunk type from `r` to `reproducibleR` for all code chunks that should be automatically checked for reproducibility.
In `reproducibleR` chunks, all newly declared variables are identified and there contents are stored in a meta data file. 
Furthermore, the content of the code chunks itself (that is, the code syntax) is also fingerprinted and stored, such that later non-reproduction can be traced back to either changes of the code chunk  or changes of the computation.
Once a document is reproduced, in R-Markdown speak _knitted_ again [@yihui2015], all computational results generated in the `reproducibleR` chunks are automatically tested for reproducibility.
A simple way to assess reproducibility of computational results would be  a bitwise comparison of the entire knitted document (i.e., the resulting HTML, PDF or DOCX file).
Yet, a particular advantage of using `reproducibleR` chunks is that users have a fine-grained control over which stages of a data analysis or any other computation should be checked for reproducibility. 
For example, if there is non-reproducibility, is it due to changes in the data file, during data cleaning or statistical modeling. 
Using `reproducibleRchunks`, users can separately assess the reproducibility of each stage of a given data analysis (or any other computation).
In the following, we will briefly demonstrate how successful and failed reproduction will be displayed, we will go into detail of the mechanics of the package briefly and discuss some limitations of the approach.

@@

The functionality of the package is to provide a `reproducibleR` chunk, which are code chunks in R Markdown documents, which can be used like regular `R`-chunks (including name labels and the usual options regulating output parameters). 
The package should be loaded in the first regular `R` code chunk in the document using
```{r eval=FALSE}
library(reproducibleRchunks)
```

The package registers a new type of code chunk called `reproducibleR`.
The following example shows a snippet from an R Markdown file, in which there is a `reproducibleR` chunk with the name _addition_. 
In the chunk, a new variable `my_sum` is declared and defined to be the sum of a variable `x` plus one. 
Variable `x` was declared in an earlier code chunk (not shown in the following screenshot) and is thus not subject to the reproducibility test of this chunk.

![](img/rstudio-screenshot-marker2.png)
In this example, the R Markdown document is knitted into a HTML-report.
On document generation, the R code in the `reproducibleR`-chunk is executed and the value of `my_sum` is stored in a separate file. 
A reproducibility report is shown below the code chunk to indicate all variables of the given chunk whose data are stored:

@@

The package executes reproducibleR code chunks as regular R code.
Once code execution is completed, it gathers information about all variables that were newly declared in the current chunk. 
The contents of those variables are stored in a separate JSON data file.
The name of the JSON file contains the original Markdown file and the chunk label and always starts with the prefix `repro`). 
That is, if reproducibility information of a chunk labelled `datacleaning` in the file `sem_analysis.Rmd` is stored in a file called _.repro_sem_analysis.Rmd_datacleaning.json_.
Once the document is re-generated and JSON data files exist, their content is checked against the newly computed chunk variables for identity.

```{r echo=FALSE}
set.seed(42)
numbers_ex <- rnorm(5)
```
Here is an example of how the contents are stored, which is a single variable called `numbers` with a vector of five numbers `r numbers_ex` generated by the following command:

```{r echo=TRUE, eval=FALSE}
numbers_ex <- rnorm(5)
```

This vector is serialized to a JSON file, which is an open and transparent format for storing contents that are both machine-readable but also human-readable:

@@

@@

![](img/schema-json-fingerprints.png)

Once a document is (re-)generated with reproducibility data present (i.e., reproducibleRchunks with matching JSON files), the package will compare all objects amenable for reproducibility checks.
Identity of original and reproduced results are checked with `identical()` function from R's `base` package, which according to the R documentation is a "safe and reliable way to test two objects (even complex ones) for being exactly identical" (see `?identical` R documentation).
Using exact comparisons of object identity is sometimes more restrictive than one would expect.
This is particularly true for numeric comparisons. 
For example results of numeric computations may be logically equivalent but not bitwise equivalent; in numeric representations, there is sometimes a bit reserved for the sign of a number, thus it can be possible to have -0 and +0, which both are exactly zero in a decimal representation but the binary representations differ.
Numeric representations may also differ up to a given numeric precision, for example, when comparing reproducibility across machines that work with different numerical precision (32bit vs 64bit). To avoid such problems, `reproducibleRchunks` rounds numeric values to a given precision (by default, eight digits)
```{r echo=FALSE, eval=TRUE, results="asis"}
jsonlite::serializeJSON(digest::digest(numbers_ex))
```

@@

```{reproducibleR abc}
xu <- rnorm(10, mean=0, sd=1) + e
y <- 4 * 4
```

@@

```{r echo=TRUE, eval=FALSE}
    options(reproducibleRchunks.templates = list(
      html="<div style='border: 3px solid black; padding: 10px 10px 10px 10px; background-color: #EEEEEE;'><h5>${title}</h5>${content}</div>"))
```

@@

- Error `In get_engine(options$engine) :
  Unknown language engine 'reproducibleR' This error is shown if a R Markdown file is knitted and the `reproducibleRchunks` package was not loaded. Fix this by adding the statement `library(reproducibleRchunks)` in the first code chunk of the R Markdown document.

@@

We strongly discourage making reproducibility checks on complete raw data. 
This is mainly for two reasons.
First, there may arise unwanted data protection issues when raw data is (accidentally) stored, copied or published via the reproducibility JSON files.
Second, the JSON format is inefficient.
It is designed to be human-readable and conformable with versioning tools such as git.
It is not meant for storing large data and will blow up file size by a considerable factor (while at the same time losing numerical precision).
At the same time, ensuring that the data set identical to the one used in the original analysis was used, adding a check on the contents of a `data.frame

@@

Even though, a variety of approaches have been suggested to ensure reproducibility of computations in R Markdown documents [], we ourselves have encountered various situations, in which such approaches (including our own) failed. 
First and foremost, many approaches rely on further software packages such as Docker, which essentially provide virtual environments that run identical on different machines (or the same machine at different time points, e.g., before and after upgrade of R or some or all packages used).
On some machines, Docker may simply be not available, either because it is not (yet) available on a certain newly introduced hardware (e.g., when Apple switched to their own processors) or because a user does not have admin privileges on a machine.
Further some approaches rely on service providers such as the MRAN archive, which was unexpectedly terminated. 
These different Achilles' heels at least occasionally lead to situations, in which users will try to locally reproduce historic computational results.
With our package, the success of these reproductions becomes easily testable.

>>>>

---
title             : "`reprodubileRchunks`: R Markdown Code Chunks with Automated Reproducibility Tests"
shorttitle        : "Title"

@@

abstract: |
 Computational results are considered _reproducible_ if the same computation on the same data yields the same results if performed on a different computer or on the same computer later in time. Reproducibility is a prerequisite for replicable, robust and transparent research in digital environments.
 Various approaches have been suggested to increase chances of reproducibility.
 Many of them rely on R Markdown as a language to dynamically generate reproducible research assets (e.g., reports, posters, presentation).
 However, a simple way to test whether different parts of these assets are reproducible is still missing.
 We introduce the R package `reproducibleRchunks`, which provides a new type of code chunk in R Markdown documents, which automatically store meta data about original computational results and check later reproductions of the research assets for reproduction. 
 With a minimal change to users' workflows, we hope that this approach increases transparency and trustworthiness of research assets.

keywords          : "reproducibility, Markdown, R"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"

@@

Computational results, including those from statistical data analyses, are considered _reproducible_ if the same computation on the same data yields the same results if performed on a different computer or on the same computer later in time. Reproducibility is a prerequisite for replicable, robust and transparent research in digital environments.
 Various approaches have been suggested to increase chances of reproducibility [@peikert2021reproducible, @renv2022, @chan2023rang, @nagraj2023pracpac].
 Many approaches rely on R Markdown.
R Markdown is a simple language, which allows mixing natural language, simple formatting instructions (e.g., what is a headline, what should appear in bold face, what are list elements) and computer code (R, python or others) in a single document [@xie2018r].
When ever the document is compiled (or, _knitted_ in R speak), the computer code chunks are executed and allow for the dynamic generation of parts of the document.
A particular advantage of R Markdown is that it helps avoiding common threats to reproducibility of code, first and foremost, copy&paste errors between results of a regular R script and a scientific report (e.g., a Word document) [@peikert2021reproducible];
further, it is suitable to generate a variety of scientific assets such as posters [@posterdown2019], presentations [@xaringan2022], resumés [@vitae2023] or preregistrations [@peikert2021reproducible] from the same Markdown language.

The `reproducibleRchunk` package allows users to make computational results in R Markdown documents automatically testable for reproduction (does the same script with the same data produce the same results, e.g. on a different computer and/or later in time) with minimal changes to your workflow.
There is only a single thing users need to change in their Markdown document if they are already using R Markdown: In the first code chunk of their document, load the package and then change the code chunk type from `r` to `reproducibleR` for all code chunks that should be automatically checked for reproducibility.
In `reproducibleR` chunks, all newly declared variables are automatically identified and their contents are stored in a meta data file. 
Furthermore, the code chunk itself (that is, the code syntax) is also fingerprinted and stored, such that later non-reproduction can be traced back to either changes of the code chunk or changes of the computation.
Once a document is reproduced, in R-Markdown-speak _knitted_ again [@yihui2015], all computational results generated in the `reproducibleR` chunks are automatically tested for reproducibility.
The result of the reproduction attempt is displayed in a reproducibility report for each chunk.
Customizations of these reports are possible as we will demonstrate later.
A simple way to assess reproducibility of computational results would be  a bitwise comparison of the entire knitted document (i.e., the resulting HTML, PDF or DOCX file).
Yet, a particular advantage of using `reproducibleR` chunks is that users have a fine-grained control over which stages of a data analysis or any other computation should be checked for reproducibility. 
For example, if there is non-reproducibility, is it due to changes in the data file, during data cleaning or statistical modeling. 
Using `reproducibleRchunks`, users can separately assess the reproducibility of each stage of a given data analysis (or any other computation).
Also, not all elements of a research asset are required to reproduce (such as the current date displayed in a presentation of results).
In the following, we will briefly demonstrate how successful and failed reproduction will be checked and displayed by the package, we will go into details of the mechanics of the package, and discuss potential cases of non-reproducibility and how the package will handle these.

@@

The core functionality of the package is to provide a `reproducibleR` chunk, which are code chunks in R Markdown documents that can be used like regular `R`-chunks (including name labels and the usual options regulating output parameters). 
The package should be loaded in the first regular `R` code chunk in the document using
```{r eval=FALSE, echo=TRUE}
library(reproducibleRchunks)
```
On load, the package registers a new type of code chunk called `reproducibleR`.
The following example shows a snippet from an R Markdown file, in which there is a `reproducibleR` chunk with the name _addition_. 
In the chunk, a new variable `my_sum` is declared and defined to be the sum of a variable `x` plus one. 
Variable `x` was declared in an earlier code chunk (not shown in the following screenshot) and is thus not subject to the reproducibility test of this chunk.

![](img/rstudio-screenshot-marker2.png)
In this example, the R Markdown document is knitted into a HTML-report.
On document generation, the R code in the `reproducibleR`-chunk is executed and the value of `my_sum` is stored in a separate file. 
A reproducibility report is shown below the code chunk to indicate all variables of the given chunk whose data are stored. The following is a snippet from the HTML generated on the first ever creation of the document:

@@

In the following, we will describe how the package stores, retrieves and compares reproducibility information.
First of all, the package executes `reproducibleR` code chunks just as regular R code chunks.
That is, there are no additional limits imposed on what can be computed and tested for reproducibility.
Once code execution is completed, the package gathers information about all variables that were newly declared in the current chunk. 
The contents of those variables are stored in a separate JSON data file.
JSON is a shorthand for Java Script Object Notation and its main features are that it is an open standard for mapping complex objects to text files with high simplicity and readability for machines and humans [@lennon2009introduction].
The name of the JSON file contains the original Markdown file and the chunk label and always starts with the prefix `repro`). 
That is, if reproducibility information of a chunk labelled `datacleaning` in the file `sem_analysis.Rmd` is stored in a file called _.repro_sem_analysis.Rmd_datacleaning.json_.
Once the document is re-generated and JSON data files exist, their content is checked against the newly computed chunk variables for identity.

```{r echo=FALSE, eval=TRUE}
set.seed(42)
numbers_ex <- sample(1:10, 5)
```
Here is an example of how the contents of a single variable called `numbers` is stored in JSON format. Here, the variable content is a vector of five random draws from the number one to ten: `r numbers_ex` generated by the following command:

```{r echo=TRUE, eval=FALSE}
numbers_ex <- sample(1:10, 5)
```

This vector can be serialized in raw format to a JSON format as follows:

@@

For privacy reasons and to save disk space, we actually do not store the raw data by default but only fingerprints of the data.
Fingerprints are realized via so-called one-way hash functions, which can take an arbitrary large digital object and map it onto a fixed-size object, which is typically displayed as hexadecimal string.
Fingerprints allow for an efficient comparison of identity of results while avoid data security and privacy issues as they do not allow to reproduce the original data.
The following shows how the information about variable `numbers` is stored using a SHA256-fingerprint:

```{r echo=FALSE, eval=TRUE, results="asis"}
jsonlite::serializeJSON(digest::digest(numbers_ex,algo = "sha256"))
```

In addition to the contents of all newly declared variables in a `reproducibleR`-chunk, the package also stores fingerprints of the code syntax.
This is shown in the following schematic figure:
![](img/schema-json-fingerprints.png)

@@

Once a document is (re-)generated with reproducibility data present (i.e., reproducibleRchunks with matching JSON files), the package will compare all objects amenable for reproducibility checks.
Identity of original and reproduced results are checked with `identical()` function from R's `base` package, which according to the R documentation is a "safe and reliable way to test two objects (even complex ones) for being exactly identical" (see `?identical` R documentation).
Using exact comparisons of object identity is sometimes more restrictive than one would expect.
This is particularly true for numeric comparisons. 
For example results of numeric computations may be logically equivalent but not bitwise equivalent; in numeric representations, there is sometimes a bit reserved for the sign of a number, thus it can be possible to have -0 and +0, which both are exactly zero in a decimal representation but the binary representations differ.
Numeric representations may also differ up to a given numeric precision, for example, when comparing reproducibility across machines that work with different numerical precision (32bit vs 64bit). To avoid such problems, `reproducibleRchunks` rounds numeric values to a given precision (by default, eight digits)

@@

```{reproducibleR abc,echo=TRUE,eval=TRUE}
xu <- rnorm(10, mean=0, sd=1) + e
y <- 4 * 4
```

@@

```{r echo=TRUE, eval=FALSE}
    options(reproducibleRchunks.templates = list(
      html="<div style='border: 3px solid black; 
      padding: 10px 10px 10px 10px; 
      background-color: #EEEEEE;'>
      <h5>${title}</h5>${content}</div>"))
```

@@

- Error `In get_engine(options$engine) :  Unknown language engine 'reproducibleR'` This error is shown if a R Markdown file is knitted and the `reproducibleRchunks` package was not loaded. Fix this by adding the statement `library(reproducibleRchunks)` in the first code chunk of the R Markdown document.

@@

We strongly discourage making reproducibility checks on complete raw data. 
This is mainly for two reasons.
First, there may arise unwanted data protection issues when raw data is (accidentally) stored, copied or published via the reproducibility JSON files.
Second, the JSON format is inefficient in terms of storage space required.
It is designed to be human-readable and conformable with versioning tools such as git.
It is not meant for storing large data and will blow up file size by a considerable factor (while at the same time losing numerical precision).
At the same time, ensuring that the data set identical to the one used in the original analysis was used, adding a check on the contents of a `data.frame`.

@@

Even though, a variety of approaches have been suggested to ensure reproducibility of computations in R Markdown documents [@peikert2021reproducible, @renv2022, @chan2023rang, @nagraj2023pracpac], we ourselves have encountered various situations, in which such approaches (including our own) failed. 
First and foremost, many approaches rely on further software packages such as Docker, which essentially provide virtual environments that run identical on different machines (or the same machine at different time points, e.g., before and after upgrade of R or some or all packages used).
On some machines, Docker may simply be not available, either because it is not (yet) available on a certain newly introduced hardware (e.g., when Apple switched to their own processors) or because a user does not have admin privileges on a machine.
Further some approaches rely on service providers such as the MRAN archive, which was unexpectedly terminated. 
These different Achilles' heels at least occasionally lead to situations, in which users will try to locally reproduce historic computational results.
With our package, the success of these reproductions becomes easily testable.
